---
title: "MRA - Assignment 2"
format: pdf
editor: visual
---

## Q3. Model Selection

```{r}
library(tidyverse)
library(dplyr)
library(olsrr)
```

```{r}
crime_data = read_csv("Crimes.csv")
crime_data
```

#### a) Use R to fit all possible models and compute AIC, BIC and R2 adj for each model. Report a table with your results.

```{r}
crime_model = lm(VR ~ ., data = crime_data)
tab = ols_step_all_possible(crime_model)
results = tab$result[,c("predictors","aic","sbc","adjr")]
results
```

#### b) Indicate the best model overall according to each of AIC, BIC and R2 adj.

```{r}
print(c(results$predictors[which.min(results$aic)],results$aic[which.min(results$aic)]))
print(c(results$predictors[which.min(results$sbc)],results$sbc[which.min(results$sbc)]))
print(c(results$predictors[which.max(results$adjr)],results$adjr[which.max(results$adjr)]))
```

#### c) Implement a forward stepwise regression that uses BIC. You will start by fitting the null model (the model with no covariates) and computing its BIC. Then, consider all possible one covariate models and compute their BICs. Iterate until there is no improvement in your criteria.

```{r}
combinations = list(("1"),c("P","M","MR"),c("M + P","MR + P","MR + M"),c("MR + M + P"))
min = 100000000
min_combination = ""

for(i in 1:length(combinations)){
  flag = FALSE
  for(j in 1:length(combinations[[i]])){
    f = as.formula(paste("VR ~ ",paste(combinations[[i]][j])))
    crime_model = lm(f,data = crime_data)
    lm = sum(log(dnorm(crime_data$VR,fitted.values(crime_model),
                       sd = summary(crime_model)$sigma)))
    bic = -2 * lm + (i+1) * log(nrow(crime_data))
    if(bic < min){
      min = bic
      min_combination = combinations[[i]][j]
      flag = TRUE
    }
  }
  if(flag == FALSE){
    break
  }
}

print(min)
print(min_combination)
```

#### d) Does the forward stepwise method find the best possible subset? Compare the solutions to item (a) and (c). Explain why solutions from stepwise regression might differ from the all possible regressions method in (a).

By looking at the BIC values, the best possible subset found using the all possible regressions was MR + M with a BIC value of 679.282922. The forward selection algorithm using the BIC found the same subset MR + M with a BIC value of 679.3748.

**...write more**

#### e) Explain how you could implement a backward selection algorithm using the F test as a decision rule. You can assume that the confidence level is α = 0.05.

**...write**

## Q4) Multiple regression

```{r}
football = read_csv("football.csv")
football
```

#### a)

```{r}
football_model = lm(y ~ x2 + x7 + x8,data=football)
summary(football_model)
```

#### b) ...write stuff

```{r}
y_bar = mean(football$y)

sst = sum((football$y - y_bar)^2)
sse = sum((football$y - fitted.values(football_model))^2)
ssr = sum((fitted.values(football_model) - y_bar)^2)

cat("sse = ",sse," , ","ssr = ",ssr," , ","sst = ",sst)
```

The relationship between SSt, SSe, and SSr:

SSt = SSe + SSr.

```{r}
sse + ssr
```

```{r}
p = 3
n = nrow(football)
df_sse = n - p - 1
df_ssr = p
df_sst = df_sse + df_ssr

cat("DF of sse = ",df_sse," , ","DF of ssr = ",df_ssr," , ","DF of sst = ",df_sst)
```

```{r}
mst = sst / df_sst
mse = sse / df_sse
msr = ssr / df_ssr
cat("MSE = ",mse," , ","MSR = ",msr," , ","MST = ",mst)
```

#### c)

```{r}
X = model.matrix(football_model)
beta = solve(t(X)%*%X)%*%t(X)%*%football$y
sigma_2 = sse / (n - p - 1)
sigma = sigma_2 * solve(t(X)%*%X)

t_b1 = beta[2,1] / sqrt(sigma[2,2])
t_b2 = beta[3,1] / sqrt(sigma[3,3])
t_b3 = beta[4,1] / sqrt(sigma[4,4])

print(c("T_" = t_b1, "T_" = t_b2, "T_" = t_b3))
```

#### d)

```{r}
r2 = 1 - (sse/sst)
r2_adj = 1 - (mse/mst)

print(c(r2 = r2,r2_adj = r2_adj))
```

#### e)

```{r}
f_stat = msr/mse
c(f_stat = f_stat,"95% CI F distribution" = qf(0.95,p, n - p - 1))
```

Reject H0

#### f)

```{r}
c("Square of correlation coefficent" = cor(football$y, fitted.values(football_model))^2, "R2" = r2)
```

#### g)

```{r}
X_Star = matrix(c(1,2300,56,2100),ncol=1)

Y_Star_mean = t(X_Star)%*%matrix(beta,ncol=1)

Y_Star_mean_upper = Y_Star_mean + qt(0.95,n-p-1) * sqrt(sigma_2 * t(X_Star)%*%solve(t(X)%*%(X))%*%X_Star)
Y_Star_mean_lower = Y_Star_mean - qt(0.95,n-p-1) * sqrt(sigma_2 * t(X_Star)%*%solve(t(X)%*%(X))%*%X_Star)

c(lower = Y_Star_mean_lower,upper = Y_Star_mean_upper)
```

#### h)

```{r}
football_model_2 = lm(y ~ x7 + x8,data=football)
summary(football_model_2)
```

```{r}
y_bar_2 = mean(football$y)

sst_2 = sum((football$y - y_bar)^2)
sse_2 = sum((football$y - fitted.values(football_model_2))^2)
ssr_2 = sum((fitted.values(football_model_2) - y_bar)^2)

cat("sse = ",sse_2," , ","ssr = ",ssr_2," , ","sst = ",sst_2)
```

```{r}
p_2 = 2
n = nrow(football)
df_sse_2 = n - p_2 - 1
df_ssr_2 = p_2
df_sst_2 = df_sse_2 + df_ssr_2

cat("DF of sse = ",df_sse_2," , ","DF of ssr = ",df_ssr_2," , ","DF of sst = ",df_sst_2)
```

```{r}
mst_2 = sst_2 / df_sst_2
mse_2 = sse_2 / df_sse_2
msr_2 = ssr_2 / df_ssr_2
cat("MSE = ",mse_2," , ","MSR = ",msr_2," , ","MST = ",mst_2)
```

#### i)

```{r}
f_stat_2 = msr_2/mse_2
c(f_stat = f_stat_2,"95% CI F distribution" = qf(0.95,p_2,n -p_2 -1))
```

#### j)

```{r}
r2_2 = 1 - (sse_2/sst_2)
r2_adj_2 = 1 - (mse_2/mst_2)

print(c(r2 = r2_2,r2_adj = r2_adj_2))
```

#### k)

Y_Star_mean_upper - Y_Star_mean_lower

```{r}
X = model.matrix(football_model_2)

beta = solve(t(X)%*%X)%*%t(X)%*%football$y
sigma_2 = sse / (n - p_2 - 1)
sigma = sigma_2 * solve(t(X)%*%X)

X_Star = matrix(c(1,56,2100),ncol=1)
Y_Star_mean = t(X_Star)%*%matrix(beta,ncol=1)

Y_Star_mean_upper = Y_Star_mean + qt(0.95,n - p_2 -1) * sqrt(sigma_2 * t(X_Star)%*%solve(t(X)%*%(X))%*%X_Star)
Y_Star_mean_lower = Y_Star_mean - qt(0.95,n - p_2 - 1) * sqrt(sigma_2 * t(X_Star)%*%solve(t(X)%*%(X))%*%X_Star)

c(lower = Y_Star_mean_lower,upper = Y_Star_mean_upper)
```

#### l)

```{r}

```

## Q5)

```{r}
bike_sharing = read.csv("bikesharing.csv")
```

#### a)

```{r}
bike_sharing$mnth = as.factor(bike_sharing$mnth)
```

#### b)

```{r}
bike_model = lm(cnt ~ hum + windspeed + temp + mnth, data = bike_sharing)
summary(bike_model)
```

#### c)

```{r}
anova(bike_model)
```

#### d)

By looking at the summary chart of the model, we can see the following.

Months **with** significant difference compared to January, given Hum, Temp, and Windspeed:

**April, May, September, October, November, December.**

Months **without** significant difference compared to January, given Hum, Temp, and Windspeed

**February, March, June, July, August.**

#### e)

```{r}
X_Star = matrix(nrow = 0,ncol = 15)
for(i in 1:12){
  x = c(1,0.4,0.3,0.65,0,0,0,0,0,0,0,0,0,0,0)
  if(i!=1){
    x[3+i] = 1
  }
  X_Star = rbind(X_Star,x)
}


result = X_Star%*%matrix(coef(bike_model))
dimnames(result) = list(c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec"),c("Predicted value"))
knitr::kable(result)
```
